Hyperplane Elimination for Quickly Enumerating Local Optima

### 2015-09-24 Hyperplane Enumeration/main.tex
Incorporated Bill's edits

\begin{abstract}
Examining the properties of local optima is a common method for understanding
combinatorial-problem landscapes.
Unfortunately,
exhaustive methods for finding local optima are limited to very small problem sizes.
We propose a method for exploiting generalized problem structure
to skip hyperplanes
that cannot contain local optima. In most of our experiments,
this hyperplane elimination allows enumeration time to scale with
the total number of local optima instead of the total landscape size.
We prove optimality for linear and $k$-bounded problems, and we provide
empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
Furthermore, we are able to refine this method to find only solutions
that cannot be improved by flipping $r$ or fewer bits. Counterintuitively,
increasing $r$ can reduce total runtime by drastically reducing the
number of local optima. While previous methods were limited to
landscapes with at most $2^{30}$ binary strings, hyperplane elimination can enumerate the same problems with
$2^{77}$ binary strings, and find all 4-bit local optima to problems with $2^{200}$ binary strings.
We believe this increase in size will help further our understanding of complex real-world landscapes.

@@

\section{Mk Landscapes as a tool for Problem Generalization}
\label{sec-mk}
An Mk Landscape~\cite{whitley:2015:mk} is any function
who's value is equal to the sum of a set of subfunctions.
Each subfunction uses a small subset of variables from the original function's input.
Combined with limitations on the number of subfunctions and the size of each subfunction's
subset, Mk Landscapes can be efficiently searched for both local~\cite{whitley:2012:constant,chicano:2014:ball}
and global~\cite{goldman:2015:GBO,tintos:2015:partitioncross} optima.

@@

The Concatenated Trap problem~\cite{deb:1992:trap} is a composition of $k$-order deceptive,
separable subfunctions. In Mk Landscape terms, $M=N/k$ such that $\forall_i |s_i| = k$ and
$\forall_{i \neq j} s_i \cap s_j = \emptyset$. Each $f_i$ applies an identical subfunction based
on the number of variables set to 1:
\begin{equation}
   trap(t) = \left\{
     \begin{array}{rl}
       k-1-t, &  t<k\\
       k,   &  t = k
     \end{array}
   \right.
  \label{eq-trap}
\end{equation}
While this problem is still in common use~\cite{hsu:2015:dsmgaII,inoue:2015:adaptivep3},
most advanced methods can solve it trivially~\cite{goldman:2012:ltga} and when expressed
as an Mk Landscape it can be solved exactly in \BigO{N} time~\cite{whitley:2015:mk}.
We therefore include it only because its structure allows for straight forward algorithm analysis.
Here we set $k=5$.

@@

Our final examined problem is randomly generated maximum satisfiability or MAX-kSAT.
This version of the canonical NP-Complete boolean satisfiability problem is formulated
as the maximization of $M=4.27N$ clauses, each containing exactly $k=3$ unique literals.
A clause is satisfied if any of its literals match how a solution's variables are set.

@@

These advances can be applied directly to the task of finding all local optima in a
landscape~\cite{ochoa:2015:crossovernetworks}.
Instead of requiring \BigO{N} time to evaluate each solution and to check if it
is a local optima, only \BigO{1} time is needed. Consider an enumeration ordering of the landscape
that uses gray-codes, meaning that each transition between solutions requires exactly 1 bit flip.
In Gray-Box optimization updating fitness and the list of improving moves after a single bit flip
takes \BigO{1} time. That property holds even when looking for $r$-bit fitness-improving moves.
Therefore Gray-Box enumeration is able to find all local optima in \BigO{2^N} time.

\section{Hyperplane Elimination}
\begin{figure}
  \centering
  \includegraphicsfit{Enumerate}
  \caption{Example change of enumeration ordering. The gray variables represent all dependencies
           for move $m_i$ which flips variable $F$. By reordering, $m_i$'s lowest $index$ dependency improves from 2 to 4.}
  \label{fig-enumerate}
\end{figure}

Due to the limited non-linearity of the Gray-Box domain, it is possible to exclude large
parts of the search space without missing any local optima.
Consider the representation presented in the top of Figure~\ref{fig-enumerate}. In a Black-Box
domain, enumeration would progress as a binary counter, treating $index$ zero (symbol $A$ in the solution) as
the least significant bit. This enumeration ordering ensures that before changing $index$ $i$, all possible settings of $index$
0 through $i-1$ have been tested. This corresponds to examining the hyperplane where the lowest $i-1$ positions
can vary and all other positions remain fixed. The Gray-Box domain makes it possible to skip hyperplanes
that cannot contain local optima. In Figure~\ref{fig-enumerate}, move $m_i$ which flips variable $F$ is a fitness improvement
when enumeration starts (all variables set to 0). Due to the known relationships between variables,
we know that the quality of $m_i$ only depends on variables $C$, $E$, $F$, and $H$.
Therefore, until one of those four variables are modified, the solution cannot be a local optimum.
As a result, the hyperplane **0*00*0 cannot contain a local optima and can
therefore be eliminated from consideration during enumeration.
More generally, if at any point during enumeration
there exists a fitness-improving move, no local optima can exist until at least one
dependency of that move is modified. Any hyperplane that has all of a fitness-improving
move's dependencies fixed cannot contain any local optima.

@@

This knowledge can be exploited to skip parts of the enumeration,
as shown in Algorithm~\ref{alg-enumerate}.
Before starting, each fitness-improving move is put into a table $move\_bin$
based on that move's lowest $index$ dependency. This is the first $index$ that
can be modified by enumeration that can potentially change the fitness effect of making that move.
In order to determine how much of the enumeration can be skipped, we must find
the highest $index$ in $move\_bin$ that contains a fitness-improving move,
as done by Line~\ref{alg-enumerate-bincheck}. If no move is fitness-improving,
then a local optimum has been found.

@@

When performing subsequent iterations, not all bins need to be checked. Instead, the highest $index$
bin that must be tested is the highest $index$ flipped by the previous iteration. This
simplification is possible because
the previous iterations have verified that all moves in higher $index$ bins are not fitness improving, and no action performed during
that iteration can make them fitness improving.
Furthermore, $index$ is always the location of the least significant 1 in $solution$, meaning
iteration can continue immediately from the found $index$. This is true by construction.
Initially $solution$ contains all 0s. Each time a 1 is inserted its position is equal to $index$ and
$index$ is only increased by carry operations which reset a position to 0 before increasing $index$.

@@

\section{Improved Enumeration Ordering}
\label{sec-reorder}
As a final efficiency, the order in which variables in $solution$ are $index$ed can be reordered.
When a move is fitness-improving,
the amount of search space that is skipped depends on how high its lowest $index$ dependency is. Therefore,
by rearranging the order to make its lowest $index$ dependency higher, more search space can be skipped. 
Figure~\ref{fig-enumerate} shows how changing the $index$ order of variables
improves $m_i$'s lowest $index$ dependency from 2 to 4. Consider that before reordering $m_i$ allows the hyperplane
**000000 to be skipped by Algorithm~\ref{alg-enumerate}, while after reordering the hyperplane that can be skipped is ****0000.
Now, whenever $m_i$ is a fitness improvement, Line~\ref{alg-enumerate-bincheck} in Algorithm~\ref{alg-enumerate}
skips 4 times as many solutions.

@@

\subsection{All Optimal Orderings are Greedy}
The quality of an enumeration ordering is determined by how many solutions
Algorithm~\ref{alg-enumerate} can skip when using that ordering.
Each time the loop on Line~\ref{alg-enumerate-bincheck} stops with $index>0$,
$2^{index}$ solutions have been skipped. For practical purposes we will
assume that all moves are equally likely to be fitness improving.

\begin{lemma}
All optimal enumeration orderings are orderings of moves, such that
all remaining dependencies of each move are sequentially assigned the
highest remaining position in the enumeration ordering.
\end{lemma}

\begin{proof}
Any enumeration ordering which cannot be described by an ordering of moves
must contain the following property. There must exist a position $j$
such that $x_j$ is not a dependency of any move with
a minimum dependency of $i$ or higher where $i$ is also the minimum dependency of a move $m$.
In this situation swapping $x_i$ and $x_j$ in enumeration ordering will raise $m$'s
minimum dependency by at least one, but it will not lower the minimum dependency of any
move. Therefore any ordering which contains this property cannot be optimal.
\end{proof}

@@

\begin{theorem}
All optimal enumeration orderings are greedy orderings of moves.
\end{theorem}

@@

\begin{theorem}
Not all greedy orderings are optimal enumeration orderings.
\end{theorem}

@@

We have not been able to determine a method for finding the optimal move ordering efficiently.
Therefore we will use the greedy method as it is probabilistically optimal.

@@

\begin{proof}
While there is no restriction on how variables are ordered in the problem,
performing reordering will ensure that
all variables which appear in the same $s_i$ are consecutively placed during enumeration.
As such it is possible to create a numbering of $f_i$ such that if $i < j$ then all of $s_i$'s
variables appear before $s_j$'s in enumeration ordering.

@@

Once the first local optimum is found, Algorithm~\ref{alg-enumerate} proceeds
by finding all local optima in larger and larger hyperplanes. Initially $f_0$
is enumerated to find all local optima in the hyperplane with all $f_i, i>0$
fixed. This process ends when a carry operation modifies a bit of $f_1$. At
this point $f_1$ is enumerated, such that each time $f_1$ contains no improving moves
$f_0$ is enumerated again. This finds all local optima in the hyperplane with $f_i, i>1$
fixed. We represent the time required to find all local optima in the hyperplane
with $f_j, j>i$ fixed as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated.
$T(i) = |l_i|*T(i-1)+2^k$ where $|l_i|$ is the number of ways $f_i$ can be set such
that it contains no fitness-improving moves.
Each setting in $l_i$ exposes a new hyperplane
that can contain local optima, which causes the recursive call to $T(i-1)$.
As all $|s_i| \leq k$ and $k$ is a
constant, $\forall_i |l_i| \leq c \leq 2^k$ meaning that $c$ is a constant.
The time required to find all local optima is therefore
$T(M-1)\leq \sum_0^{M-1}2^kc^i<2^kc^{M}$.
The number of local optima in the landscape is \BigO{c^{M}} meaning that Algorithm~\ref{alg-enumerate}
is within a constant factor ($2^k$) of optimal.
\end{proof}

@@

\subsubsection{Importance of Reordering}
The order in which variables are enumerated for $k$-bound separable problems
has a significant impact on Algorithm~\ref{alg-enumerate}'s complexity.
Instead of using the normal reordering procedure, consider an ordering
such that if $i < j$, all variables in $s_i$ appear before the second variable
in $s_j$ and after the first variable in $s_j$.

@@

Each algorithm was given a maximum of four hours on a 2.5GHz Intel Xeon E5-2670v2 processor to find
all $r$-bit local optima. Any run which failed was rerun once to prevent cluster issues from introducing
noise. For the same reason any run which was over ten times slower than the next slowest run of the
same configuration with a different seed was rerun. This resulted in 33 out of 36,000 runs being rerun.
If the algorithm failed to complete in the alloted time for an instance,
we declared it unsuccessful for that combination of $r$ and $N$ of that problem. All results
we report are for successful combinations. For timing purposes local optima were counted but not
recorded. All of our code, data, and statistical analysis can be downloaded from our
website.\footnote{\url{https://name.of.site/removed/to/be/anonymous/}}

\subsection{Finding Local Optima}
\label{sec-1-bit-optima}
Our first experiment is to determine the empirical complexity of each method with respect to problem size.
Of the 681 configurations where at least one method was successful, Hyper-Reorder had the lowest
mean time to find all local optima all but 2 times. In both of those cases Hyper-Reorder was within a tenth of a second
of the best method.
Figure~\ref{fig-boxplot-method} shows the variance in runtime for the largest problem size all three methods
were successful. In all cases it is clear that Hyper-Reorder outperforms the other methods.

@@

More critically, Figure~\ref{fig-length-method} shows how each method scales with problem size.
Not only is Hyper-Reorder out performing the alternatives on all sizes, the rate at which it
scales to problem complexity is better. Table~\ref{table-scaling} shows the model fit for each
method on each problem. This value is the slope of each line shown in Figure~\ref{fig-length-method}.
As expected Gray-Box scales at almost exactly $2^N$. Using hyperplane elimination without
reordering reduces this complexity, with reordering making further improvements. In the final
column we show the growth rate of the number of local optima. It is impossible for any method
which enumerates all local optima to grow slower than this rate. On Concatenated Traps, Adjacent
NKq, and Ising Spin Glass, Hyper-Reorder has an estimated slope better than optimal. This is
due to lower order terms dominating runtime for small problem sizes, as can be see in
Figure~\ref{fig-length-method}. For instance, fitting
the curve only to problems where $N>60$ makes Hyper-Reorder have $m=0.199$ on Concatenated Traps and
$m=0.358$ on Adjacent NKq. For Concatenated Traps this matches our theoretical predictions that
Hyper-Reorder scales at $2^k2^{N/k}$ as $\frac{1}{k}=0.2$.

This reduction in growth complexity translates into a significant increase in the size of problems
which can be enumerated. For instance, using the previous best enumeration methods~\cite{ochoa:2015:crossovernetworks}
the largest Adjacent NKq problem size which can be enumerated in 4 hours is 34 bits. Given the same time
Hyper-Reorder can enumerate problems with 77 bits. Using the regression model, we estimate solving
a 77 bit problem using Gray-Box would require 900~million years. Similarly, Hyper-Reorder's largest
Random NKq size of 69 bits would require Gray-Box 3~million years. These larger sizes make for
much better approximations of landscapes interesting to search methods.

@@

Perhaps most striking in Figure~\ref{fig-slope-method} is that on Adjacent NKq, Random NKq,
and Ising Spin Glass, increasing $r$ \emph{reduces} Hyper-Reorder's growth complexity.
Increasing $r$ creates more moves which can be fitness improving. As a result there
are more hyperplanes which can potentially be detected and eliminated.
On Concatenated Traps no flip using $1 < r < k$ bits can be fitness improving
without a 1-bit flip being fitness improving, meaning no additional hyperplanes can be skipped.
For MAX-kSAT, increasing $r$ likely compounds the high dependency problems discussed in
Section~\ref{sec-1-bit-optima}.

@@

Previous work~\cite{ochoa:2015:crossovernetworks} noted that counter-intuitively Adjacent
NKq has more local optima than Random NKq for $r=1$ and $N=20$, even though the former is generally
easier to solve. However, many of these optima are ``weak'' in the sense that they are not also
2-bit local optima. Here we extend those observations for up to $N=59$ and up to $r=4$. Adjacent
NKq tends to have more 1-bit local optima than Random NKq, most of which disappear as $r$
increases. Furthermore, the growth rate of local optima between the two landscapes diverges
as $r$ increases.

@@

In our original experiments we tested problems up to size $N=100$, which was sufficient to
cause all three methods using $r=1$ to fail all problems except for Hyper-Reorder on Concatenated
Traps. However, using higher $r$ values, Hyper-Reorder is able to significantly extend its range.
On Random NKq, Hyper-Reorder was successful at finding 2-bit local optima for $N=73$. On Adjacent
NKq using all $r$ tested except 1 were successful on $N=100$. To push Algorithm~\ref{alg-enumerate}
to the extreme we tested Adjacent NKq with $N=200$ and $r=4$, with all 30 runs successful. On
average these runs required 5.6 minutes to complete. However, there was significant variance
in both the runtime and number of local optima. Eighteen runs completed in under 1 minute,
while the slowest run took 1.5 hours. Three had fewer than 50 4-bit local optima, while five
had over 100,000.

\section{Conclusions and Future Work}
By expressing a function as an Mk Landscapes, we have developed a method to find all
local optima of that function very efficiently. The structure imposed by
an Mk Landscape allows for the identification of hyperplanes in the search space
which cannot contain any local optima. Combined with a greedy method for
changing the order of how the space is enumerated, we were able to show
this method's runtime complexity scales with the number of local optima, unlike
previous methods which scale with the number of possible solutions.
When applied to NKq Landscapes and Ising Spin Glasses this allowed for a dramatic
increase in the size of problems which can be enumerated. We believe that
these larger problem sizes will be more representative of the types
of search spaces where optimization is performed.

We show how this hyperplane elimination can be extended to find only solutions
which cannot be improved by flipping $r$ or fewer bits. For some problems
this counterintuitively reduces search times, and on Adjacent NKq allowed
us to find all 4-bit local optima of 200 bit problems.

>>>>

\begin{abstract}
Examining the properties of local optima is a common method for understanding
combinatorial-problem landscapes.
Unfortunately,
exhaustive methods for finding local optima are limited to very small problem sizes.
We propose a method for exploiting problem structure
to skip hyperplanes
that cannot contain local optima. In most of our experiments,
this hyperplane elimination allows enumeration time to scale with
the total number of local optima instead of with the total landscape size.
We prove optimality for linear and $k$-bounded problems, and we provide
empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
Furthermore, we are able to refine this method to find only solutions
that cannot be improved by flipping $r$ or fewer bits. Counterintuitively,
increasing $r$ can reduce total runtime by drastically reducing the
number of local optima. While previous methods were limited to
landscapes with at most $2^{30}$ binary strings, hyperplane elimination can enumerate the same problems with
$2^{77}$ binary strings, and find all 4-bit local optima to problems with $2^{200}$ binary strings.
We believe this increase in size will help further our understanding of complex real-world landscapes.

@@

\section{Mk Landscapes as a tool for Problem Generalization}
\label{sec-mk}
An Mk Landscape~\cite{whitley:2015:mk} is any function
who's value is equal to the sum of a set of subfunctions.
Each subfunction uses a small subset of variables from the original function's input.
Combined with limits on the number of subfunctions and the size of each subfunction's
subset, Mk Landscapes can be efficiently searched for both local~\cite{whitley:2012:constant,chicano:2014:ball}
and global~\cite{goldman:2015:GBO,tintos:2015:partitioncross} optima.

@@

The Concatenated Trap problem~\cite{deb:1992:trap} is a composition of $k$-order deceptive,
separable subfunctions. In Mk Landscape terms, $M=N/k$ such that $\forall_i |s_i| = k$ and
$\forall_{i \neq j} s_i \cap s_j = \emptyset$. Each $f_i$ applies an identical subfunction based
on the number of variables set to 1:
\begin{equation}
   trap(t) = \left\{
     \begin{array}{rl}
       k-1-t,~ &  t<k\\
       k,~   &  t = k
     \end{array}
   \right.
  \label{eq-trap}
\end{equation}
While this problem is still in common use~\cite{hsu:2015:dsmgaII,inoue:2015:adaptivep3},
most advanced methods can solve it trivially~\cite{goldman:2012:ltga} and when expressed
as an Mk Landscape it can be solved exactly in \BigO{N} time~\cite{whitley:2015:mk}.
We therefore include it only because its structure allows for straight forward algorithm analysis.
Here we set $k=5$.

@@

Our final problem is randomly generated maximum satisfiability or MAX-kSAT.
This version of the canonical NP-Complete boolean satisfiability problem is formulated
as the maximization of $M=4.27N$ clauses, each containing exactly $k=3$ unique literals.
A clause is satisfied if any of its literals match how a solution's variables are set.

@@

These advances can be applied directly to the task of finding all local optima in a
landscape~\cite{ochoa:2015:crossovernetworks}.
Instead of requiring \BigO{N} time to evaluate each solution and to check if it
is a local optima, only \BigO{1} time is needed. Consider an enumeration of the landscape
that uses gray-codes, meaning that each transition between solutions requires exactly 1 bit flip.
In Gray-Box optimization updating fitness and the list of improving moves after a single bit flip
takes \BigO{1} time. That property holds even when looking for $r$-bit fitness-improving moves.
Therefore Gray-Box enumeration is able to find all local optima in \BigO{2^N} time.

\section{Hyperplane Elimination}
\begin{figure}
  \centering
  \includegraphicsfit{Enumerate}
  \caption{Example change of ordering. The gray variables represent all dependencies
           for move $m_i$ which flips variable $F$. By reordering, $m_i$'s lowest $index$ dependency improves from 2 to 4.}
  \label{fig-enumerate}
\end{figure}

Due to the limited non-linearity of the Gray-Box domain, it is possible to exclude large
parts of the search space without missing any local optima.
Consider the representation presented in the top of Figure~\ref{fig-enumerate}. In a Black-Box
domain, enumeration would progress as a binary counter, treating $index$ zero (symbol $A$ in the solution) as
the least significant bit. This ordering ensures that before changing $index$ $i$, all possible settings of $index$
0 through $i-1$ have been tested. This corresponds to examining the hyperplane where the lowest $i-1$ positions
can vary and all other positions remain fixed. The Gray-Box domain makes it possible to skip hyperplanes
that cannot contain local optima. In Figure~\ref{fig-enumerate}, move $m_i$ which flips variable $F$ is a fitness improvement
when enumeration starts (all variables set to 0). Due to the known relationships between variables,
we know that the quality of $m_i$ only depends on variables $C$, $E$, $F$, and $H$.
Therefore, until one of those four variables are modified, the solution cannot be a local optimum.
As a result, the hyperplane **0*00*0 cannot contain a local optima and can
therefore be eliminated from consideration during enumeration.
More generally, if at any point during enumeration
there exists a fitness-improving move, no local optima can exist until at least one
dependency of that move is modified. Any hyperplane that has all of a fitness-improving
move's dependencies fixed cannot contain any local optima.

@@

This knowledge can be exploited to skip parts of the enumeration,
as shown in Algorithm~\ref{alg-enumerate}.
Initially all variables are set to 0 and
each fitness-improving move is put into a table $move\_bin$
based on that move's lowest $index$ dependency. This is the first $index$ that
can be modified by enumeration that can potentially change the fitness effect of making that move.
In order to determine how much of the enumeration can be skipped, we must find
the highest $index$ in $move\_bin$ that contains a fitness-improving move,
as done by Line~\ref{alg-enumerate-bincheck}. If no move is fitness-improving,
then a local optimum has been found.

@@

When performing subsequent iterations, not all bins need to be checked. Instead, the highest $index$
bin that must be tested is the highest $index$ flipped by the previous iteration. This
simplification is possible because
the previous iterations have verified that all moves in higher $index$ bins are not fitness improving, and no action performed during
that iteration can make them fitness improving.
Furthermore, $index$ is always the location of the least significant 1 bit in $solution$, meaning
iteration can continue immediately from the found $index$. This is true by construction.
Initially $solution$ contains all 0s. Each time a 1 is inserted its position is equal to $index$ and
$index$ is only increased by carry operations which reset a position to 0 before increasing $index$.

@@

\section{Reordering Variables}
\label{sec-reorder}
As a final efficiency, the order in which variables in $solution$ are $index$ed can be reordered.
When a move is fitness-improving,
the amount of search space that is skipped depends on how high its lowest $index$ dependency is. Therefore,
by rearranging the order to make its lowest $index$ dependency higher, more search space can be skipped. 
Figure~\ref{fig-enumerate} shows how changing the $index$ order of variables
improves $m_i$'s lowest $index$ dependency from 2 to 4. Consider that before reordering $m_i$ allows the hyperplane
**000000 to be skipped by Algorithm~\ref{alg-enumerate}, while after reordering the hyperplane that can be skipped is ****0000.
Now, whenever $m_i$ is a fitness improvement, Line~\ref{alg-enumerate-bincheck} in Algorithm~\ref{alg-enumerate}
skips 4 times as many solutions.

@@

\subsection{All Optimal Orderings are Greedy}
The quality of an ordering is determined by how many solutions
Algorithm~\ref{alg-enumerate} can skip when using that ordering.
Each time the loop on Line~\ref{alg-enumerate-bincheck} stops with $index>0$,
$2^{index}$ solutions have been skipped. For practical purposes we will
assume that all moves are equally likely to be fitness improving.

\begin{lemma}
All optimal orderings of variables are orderings of moves, such that
all remaining dependencies of each move are sequentially assigned the
highest remaining position in the enumeration ordering.
\end{lemma}

\begin{proof}
Any ordering of variables which cannot be described by an ordering of moves
must contain the following property. There must exist a position $j$
such that $x_j$ is not a dependency of any move with
a minimum dependency of $i$ or higher where $i$ is also the minimum dependency of a move $m$.
In this situation swapping the order of $x_i$ and $x_j$ will raise $m$'s
minimum dependency by at least one, but it will not lower the minimum dependency of any
move. Therefore any ordering which contains this property cannot be optimal.
\end{proof}

@@

\begin{theorem}
All optimal orderings of variables are greedy orderings of moves.
\end{theorem}

@@

\begin{theorem}
Not all greedy orderings are optimal orderings.
\end{theorem}

@@

We believe finding the optimal ordering of moves is NP-Hard and therefore rely on
the greedy method as it is probabilistically optimal.

@@

\begin{proof}
While there is no restriction on how variables are ordered in the problem,
performing reordering will ensure that
all variables which appear in the same $s_i$ are consecutive.
As such it is possible to create a numbering of $f_i$ such that if $i < j$ then all of $s_i$'s
variables appear before $s_j$'s in enumeration ordering.

@@

Once the first local optimum is found, Algorithm~\ref{alg-enumerate} proceeds
by finding all local optima in larger and larger hyperplanes. Initially $f_0$
is enumerated to find all local optima in the hyperplane where all $f_i, i>0$ remain
fixed. This process ends when a carry operation modifies a bit of $f_1$. At
this point $f_1$ is enumerated, such that each time $f_1$ contains no improving moves
$f_0$ is enumerated again. This finds all local optima in the hyperplane where $f_i, i>1$ remain
fixed. We represent the time required to find all local optima in the hyperplane
with $f_j, j>i$ fixed as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated.
$T(i) = |l_i|*T(i-1)+2^k$ where $|l_i|$ is the number of ways $f_i$ can be set such
that it contains no fitness-improving moves.
Each setting in $l_i$ exposes a new hyperplane
that can contain local optima, which causes the recursive call to $T(i-1)$.
As all $|s_i| \leq k$ and $k$ is a
constant, $\forall_i |l_i| \leq c \leq 2^k$ meaning that $c$ is a constant.
The time required to find all local optima is therefore
$T(M-1)\leq \sum_0^{M-1}2^kc^i<2^kc^{M}$.
The number of local optima in the landscape is \BigO{c^{M}} meaning that Algorithm~\ref{alg-enumerate}
is within a constant factor ($2^k$) of optimal.
\end{proof}

@@

\subsubsection{Importance of Reordering}
The order of variables for $k$-bound separable problems
has a significant impact on Algorithm~\ref{alg-enumerate}'s complexity.
Instead of using the normal reordering procedure, consider an ordering
such that if $i < j$, all variables in $s_i$ appear before the second variable
in $s_j$ and after the first variable in $s_j$.

@@

Each algorithm was given a maximum of four hours to find all $r$-bit local optima.
Runs were distributed across a cluster using 2.5GHz Intel Xeon E5-2670v2 processors.
Any run which failed was rerun once to prevent cluster issues from introducing
noise. For the same reason any run which was over ten times slower than the next slowest run of the
same configuration with a different seed was rerun. This resulted in 33 out of 36,000 runs being rerun.
If the algorithm failed to complete in the alloted time for an instance,
we declared it unsuccessful for that combination of $r$ and $N$ of that problem. All results
we report are for successful combinations. For timing purposes local optima were counted but not
recorded. All of our code, data, and statistical analysis can be downloaded from our
website.\footnote{\url{https://name.of.site/removed/to/be/anonymous/}}

\subsection{Finding Local Optima}
\label{sec-1-bit-optima}
Our first experiment is to determine the empirical complexity of each method with respect to problem size.
We tested 932 task configurations (problem, $N$, $r$), with 681 completed by at least one method.
Of those, Hyper-Reorder had the lowest
mean time to find all local optima in all but 2. In both of those cases Hyper-Reorder was within a tenth of a second
of the best method.
Figure~\ref{fig-boxplot-method} shows the variance in runtime for the largest problem size where all three methods
were successful. In all cases it is clear that Hyper-Reorder outperforms the other methods.

@@

More critically, Figure~\ref{fig-length-method} shows how each method scales with problem size.
Not only does Hyper-Reorder outperform the alternatives on all sizes, the rate at which it
scales to problem complexity is better. Table~\ref{table-scaling} shows the model fit for each
method on each problem. This value is the slope of each line shown in Figure~\ref{fig-length-method}.
As expected, Gray-Box scales at almost exactly $2^N$. Using hyperplane elimination without
reordering reduces this complexity, and reordering makes further improvements. In the final
column we show the growth rate of the number of local optima. It is impossible for any method
which enumerates all local optima to grow slower than this rate.
However, Hyper-Reorder has an estimated slope better than optimal on Concatenated Traps, Adjacent
NKq, and Ising Spin Glass.
This is
due to lower order terms dominating runtime for small problem sizes, as can be see in
Figure~\ref{fig-length-method}. For instance, fitting
the curve only to problems where $N>60$ makes Hyper-Reorder have $m=0.199$ on Concatenated Traps and
$m=0.358$ on Adjacent NKq. For Concatenated Traps this matches our theoretical predictions that
Hyper-Reorder scales at $2^k2^{N/k}$ as $\frac{1}{k}=0.2$.

This reduction in growth complexity translates into a significant increase in the size of problems
which can be enumerated. For instance, using the previous best enumeration methods~\cite{ochoa:2015:crossovernetworks},
the largest Adjacent NKq problem size which can be enumerated in 4 hours is 34 bits. Given the same amount of time,
Hyper-Reorder can enumerate problems with 77 bits. Using the regression model, we estimate solving
a 77 bit problem using Gray-Box would require 900~million years. Similarly, Hyper-Reorder's largest
Random NKq size of 69 bits would require Gray-Box 3~million years. We believe these larger problems
are more likely to share characteristics with those where search algorithms are actually applied.

@@

Perhaps most striking in Figure~\ref{fig-slope-method} is that on Adjacent NKq, Random NKq,
and Ising Spin Glass, increasing $r$ \emph{reduces} Hyper-Reorder's growth complexity.
Increasing $r$ creates more moves which can be fitness improving. As a result there
are more hyperplanes which can potentially be detected and eliminated.
On Concatenated Traps, no flip using $1 < r < k$ bits can be fitness improving
without a 1-bit flip being fitness improving, meaning no additional hyperplanes can be skipped.
For MAX-kSAT, increasing $r$ likely compounds the high dependency problems discussed in
Section~\ref{sec-1-bit-optima}.

@@

Previous work~\cite{ochoa:2015:crossovernetworks} noted that, counter-intuitively, Adjacent
NKq has more local optima than Random NKq for $r=1$ and $N=20$, even though the former is generally
easier to solve. However, many of these optima are ``weak'' in the sense that they are not also
2-bit local optima. Here we extend those observations for up to $N=59$ and up to $r=4$. Adjacent
NKq tends to have more 1-bit local optima than Random NKq, most of which disappear as $r$
increases. Furthermore, the growth rate of local optima between the two landscapes diverges
as $r$ increases.

@@

In our original experiments we tested problems up to size $N=100$, which was sufficient to
cause all three methods using $r=1$ to fail all problems except for Hyper-Reorder on Concatenated
Traps. For Gray-Box and Hyper, increasing $r$ only decreased the largest size they could
enumerate.
However, using higher $r$ values, Hyper-Reorder is able to significantly extend its range.
On Random NKq, Hyper-Reorder was successful at finding 2-bit local optima for $N=73$. On Adjacent
NKq for all $r$ tested except 1, Hyper-Reorder was successful on $N=100$. To push Algorithm~\ref{alg-enumerate}
to the extreme, we tested Adjacent NKq with $N=200$ and $r=4$, with all 30 runs successful. On
average these runs required 5.6 minutes to complete. However, there was significant variance
in both the runtime and number of local optima. Eighteen runs completed in under 1 minute,
while the slowest run took 1.5 hours. Three had fewer than 50 4-bit local optima, while five
had over 100,000.

\section{Conclusions and Future Work}
By expressing a function as an Mk Landscapes, we have developed a method to find all
local optima of that function very efficiently. The structure imposed by
an Mk Landscape allows for the identification of hyperplanes in the search space
which cannot contain any local optima and can therefore be eliminated from enumeration.
Combined with a greedy method to
change the ordering of variables, we were able to show
this method's runtime complexity scales with the number of local optima, unlike
previous methods which scale with the number of possible solutions.
When applied to NKq Landscapes and Ising Spin Glasses, this allowed for a dramatic
increase in the size of problems which can be enumerated. We believe that
these larger problem sizes will be more representative of the types
of search spaces where optimization is performed.

We also showed how this hyperplane elimination can be refined to find only solutions
which cannot be improved by flipping $r$ or fewer bits. For some problems
this counterintuitively reduces search times, and on Adjacent NKq allowed
us to find all 4-bit local optima of 200 bit problems.
